2.3 Prob Model for PCR Positivity Estimation

### 2024-04-04 2_scripts/2.2.a CBS Preprocessing and Fig.qmd
clean the code, do the abbott~roche model

##### 2.3.1 Pre-process Canpath data
```{r}
Canpaths %>%
  select(region, assay_id) %>%
  group_by(region, assay_id) %>%
  summarise(Freq=n())

@@

#### 3.2 Canpath: make plots by region
```{r}
# Reshape meltted data
Canpaths_melt<-reshape2::melt(select(Canpaths_pr, samplemonth, region,
                                       prop.n, prop.n_RG), 
                        id.vars = c("samplemonth", "region"),
                        variable.name = "Adj",
                        value.name = "prop")

>>>>

##### 2.3.1 Pre-process Canpath data

```{r}
Canpaths %>%
  select(region, assay_id) %>%
  group_by(region, assay_id) %>%
  summarise(Freq=n())

@@

#### 3.2 Canpath: make plots by region

```{r}
# Reshape meltted data
Canpaths_melt<-reshape2::melt(select(Canpaths_pr, samplemonth, region,
                                       prop.n, prop.n_RG), 
                        id.vars = c("samplemonth", "region"),
                        variable.name = "Adj",
                        value.name = "prop")

### 2024-04-04 2_scripts/2.3 Prob Model for PCR pos.qmd
clean the code, do the abbott~roche model

load(file = "../1_data/private/RFD4682e.RData")
```

@@

## add interp var and study ID
dta_01s$study<-rep(1, length(dta_01s$pid))
dta_01s$time_pso<-rep("never_inf", length(dta_01s$pid))
rm(tb1, tb2)
```

@@

# correct var names
colnames(dta_03s)<-c("roche_n", "abbott_n",
                     colnames(dta_03s[,3:6]))
```

```{r}
## Data-04, 2/8 months data PSO, extracted from Y. Nagakama et al.
X4_Nakagama_all <- read_excel("../1_data/private/Prob Model/4-Nakagama_all.xlsx")
dta_04<-X4_Nakagama_all 
dta_04$time_pso[dta_04$time_pso=="2 months"]<-"less_3_month"
dta_04$time_pso[dta_04$time_pso=="8 months"]<-"more_3_month"

# All participants been sampled twice, wt=0.5
dta_04$sample_wt<-rep(0.5, length(dta_04$pid))

# correct col names
colnames(dta_04)<-c("study", colnames(dta_04[,2:6]))
```

```{r}
## Data-06, shared by Dr Paul Bieniasz
X6_Paul <- read_excel("../1_data/private/Prob Model/6-2024_02_06_SR1407_McGill.xlsx", 
    sheet = "data_import", skip = 1)
dta_06<-select(X6_Paul, 1:6, 9)
colnames(dta_06)<-c("age", "gender", "pid", "time",
                    "abbott_n", "roche_n", "study")
dta_06s<-dta_06[!(is.na(dta_06$abbott_n) &
                  is.na(dta_06$roche_n)),]

# add time_pso based on time
dta_06s$time_pso[dta_06s$time<=90]<-"less_3_month"
dta_06s$time_pso[dta_06s$time> 90]<-"more_3_month"
table(dta_06s$time_pso, useNA = 'ifany')

# add weights
tb1<-as.data.frame(table(dta_06s$pid))
colnames(tb1)<-c("pid","freq")
tb1$sample_wt<-1/tb1$freq
dta_06s<-merge(dta_06s, select(tb1, 1,3), by="pid",all.x = T, all.y = F)
rm(tb1, tb2)

dta_06s<-select(dta_06s, -c("time"))
```

@@

```{r}
dta_all<-dplyr::bind_rows(dta_01s, dta_03s, dta_04, dta_06s)

# add log quant assay
dta_all$log_roche_n<-log(dta_all$roche_n)    # COI for log_n = 0
dta_all$log_abbott_n<-log(dta_all$abbott_n)  # COI for log)n = 0.3365 (COI = 1.4)

# save the rbinded harmonized data
save(dta_all, file="../1_data/private/Prob Model/dta_all.RData")
```

@@

## Create a small subset for PCR+ data

```{r}
APL_PCR<-RFD4682_e %>%
  .[is.na(.$`N-IgG_RESULT`)==F & .$NAAT_STATUS %in% c("Negative","Positive"),]
APL_PCR$NAAT[APL_PCR$NAAT_STATUS=="Positive"]<-1
APL_PCR$NAAT[APL_PCR$NAAT_STATUS=="Negative"]<-0
APL_PCR$NAAT<-as.numeric(APL_PCR$NAAT)
table(APL_PCR$NAAT_STATUS, APL_PCR$NAAT, useNA = 'ifany')

# A subset for those who got results within 90 days
APL_PCR90<-APL_PCR %>%
  .[.$DAYS_SINCE_NAAT>=0 & .$DAYS_SINCE_NAAT<=90,]
```

## Build regression model using the data

```{r}
# Model 1: add age and sex as covariates, 90 days only
plm01<-glm(NAAT ~ AGE_AT_COLLECTION + GENDER  + N, data = APL_PCR90,
           family = binomial(link = "logit"))
summary(plm01)

# Model 2: add age and sex as covariates, everyone
plm02<-glm(NAAT ~ AGE_AT_COLLECTION + GENDER  + N, data = APL_PCR,
           family = binomial(link = "logit"))
summary(plm02)

# Model 3, add NAAT time into the model and compare it with model 2
plm03<-glm(NAAT ~ AGE_AT_COLLECTION + GENDER  + N + DAYS_SINCE_NAAT, data = APL_PCR,
           family = binomial(link = "logit"))
summary(plm03)

lrtest(plm02, plm03) # Sig difference, much lower AIC
```

Based on model 2 we can also add time since last PCR test into analysis, and it will improve AIC greatly for model fitting. However, this model is not applicable for other datasets as the PCR test info is missing.

```{r}
# Add predicted prob back to the data
APL_PCR90$NAAT_p<-predict(plm01, type = 'response')
hist(APL_PCR90$NAAT)

sum(APL_PCR90$NAAT)    # N=1008
sum(APL_PCR90$NAAT_p)  # N=1008
```

## Multinomial Regression Model:

```{r}
mnlm01<-multinom(relevel(as.factor(time_pso), ref = "never_inf") ~ roche_n, data = dta_all)
summary(mnlm01)

@@

# Model 02 takes sample weights into account
mnlm02<-multinom(relevel(as.factor(time_pso), ref = "never_inf") ~ roche_n, 
                 data = dta_all, weights = sample_wt)
summary(mnlm02)
```

@@

```{r}
# Model 02 takes sample weights into account
mnlm03<-multinom(relevel(as.factor(time_pso), ref = "never_inf") ~ log_roche_n, 
                 data = dta_all, weights = sample_wt)
summary(mnlm03)
```

##====================== Part II. Applying the model to APL data =============##

Apply the APL_90 model to everyone in the dataset, then plot the difference btw those who has PCR test results (observed) and those who do not (predicted).

Compared to model 1, model 2 is predicting "ever predicted" rather than "infected in past three months" as it does in model 1. The only one we can apply to the APL data is the model 1.

need to rewrite this section

#### 2.1 Prepare for the dataset

```{r}
#those tested within last 90 days
summary(APL_PCR90$DAYS_SINCE_NAAT) # Model 1 is built upon this dataset

# Add an indicator for those who had PCR test in past 90 days
RFD4682_e$PCR90<-rep(0, length(RFD4682_e$order_ID))
RFD4682_e$PCR90[RFD4682_e$order_ID %in% APL_PCR90$order_ID]<-1
table(RFD4682_e$PCR90)

# Clean the gender variable (level unknown)
table(RFD4682_e$GENDER)
APL_Clean<-RFD4682_e
APL_Clean$GENDER[APL_Clean$GENDER=="Unknown"]<-NA

# Recode NAAT as 0/1/9
APL_Clean$NAAT[APL_Clean$NAAT_STATUS=="Positive"]<-1
APL_Clean$NAAT[APL_Clean$NAAT_STATUS=="Negative"]<-0
APL_Clean$NAAT[APL_Clean$NAAT_STATUS=="No Prior NAT"]<-9
APL_Clean$NAAT<-as.numeric(APL_Clean$NAAT)
table(APL_Clean$NAAT_STATUS, APL_Clean$NAAT, useNA = 'ifany')

# Apply model 1 to APL data (full)
APL_Clean$NAAT_p<-predict(plm01, newdata = APL_Clean, type = 'response')

# Create NAAT_m, using real NAAT results for those tested <=90 days
APL_Clean$NAAT_m <- APL_Clean$NAAT_p
APL_Clean$NAAT_m[APL_Clean$PCR90==1 & APL_Clean$NAAT_STATUS=="Positive"]<-1
APL_Clean$NAAT_m[APL_Clean$PCR90==1 & APL_Clean$NAAT_STATUS=="Negative"]<-0

summary(APL_Clean$NAAT_m)
table(APL_Clean$NAAT_m==1) # number matches
table(APL_Clean$NAAT_m==0)
```

#### 2.2 Based on new data, making plot seperately

```{r}
# Calculate positivity for each month

# tb1-the data (PCR<90) used to build model
tb1<-APL_Clean[APL_Clean$PCR90==1,] %>%
  group_by(Col_Month) %>%
  summarise(n = n() , pos= mean(NAAT_m))
tb1$PCR90<-rep(1, length(tb1$Col_Month))

# tb2-we applied the model to the rest of APL (PCR>90) 
tb2<-APL_Clean[APL_Clean$PCR90==0,] %>%
  group_by(Col_Month) %>%
  summarise(n = n() , pos= mean(NAAT_m, na.rm=TRUE))
tb2$PCR90<-rep(0, length(tb2$Col_Month))

# Merge the data to take a look
tb3<-merge(tb1, tb2, by = "Col_Month", all.x = T, all.y = T)
tb3$Col_Month<-as.yearmon(tb3$Col_Month)
tb3<-tb3[order(tb3$Col_Month),]

# Compare the difference
# Paired t-test
t.test(tb3$pos.x[18:27], tb3$pos.y[18:27], paired=1)
# p-value=0.2761, no sig diff between observed (<90D) and the rest
```

#### The plots

```{r}
APL_Pmelt<-rbind.data.frame(select(tb1, Col_Month, pos, PCR90), 
                            select(tb2[18:27,], Col_Month, pos, PCR90))
APL_Pmelt$PCR90[APL_Pmelt$PCR90==0]<-"Predicted"
APL_Pmelt$PCR90[APL_Pmelt$PCR90==1]<-"Observed"

APL_Pmelt %>%
  ggplot(aes(x=Col_Month, y=pos, group=PCR90, colour=PCR90)) +
  geom_point() +
  geom_line()  +
  geom_hline(yintercept=0, colour='blue', linetype=2) +
  theme_bw() 
```

#### 2.3 Model adjusting for age and sex

@@

##============ Part III-1 Apply Model to CBS Data ================

@@

CBS_prob_r<-as.data.frame(predict(mnlm03, newdata = CBS, "probs")) # num of >3 months = 0
mean(CBS_prob_r[,1], na.rm = T)  # 59.97%
mean(CBS_prob_r[,2], na.rm = T)  # 27.70%
mean(CBS_prob_r[,3], na.rm = T)  # 12.33%

CBSs<-cbind.data.frame(select(CBS, pid, sampledate, sex, age, fsa,
                        province, roche_n, abbott_n, roche_s, log_roche_n),
                 CBS_prob_r)
```

```{r}
## Applying the model with age & sex
CBS$gender<-CBS$sex
CBS_prob_r2<-as.data.frame(predict(mnlm03.b, newdata = CBS, "probs")) # num of >3 months = 0
mean(CBS_prob_r2[,1], na.rm = T)  # 59.49%
mean(CBS_prob_r2[,2], na.rm = T)  # 32.45%
mean(CBS_prob_r2[,3], na.rm = T)  # 8.06%

@@

# Creat a ever inf indicator
CBSs$ever_inf <- 1-CBSs$never_inf

# Copy new vars to CBSs2
CBSs2$region<-CBSs$region
CBSs2$samplemonth<-CBSs$samplemonth
CBSs2$ever_inf <- 1-CBSs2$never_inf
```

@@

@@

```{r}
# Reshape meltted data
tb_melt<-reshape2::melt(select(tb_all, 1,3:5), 
                        id.vars = c("samplemonth", "region"),
                        variable.name = "inf",
                        value.name = "prop")

@@

## Part III-2 Apply model to APL data

@@

## Part III-3 Apply model to CANPATH data

>>>>

# The APL data
load(file = "../1_data/private/RFD4682e.RData")
```

@@

## add interp var and study ID
dta_01s$study<-rep(1, length(dta_01s$pid))
dta_01s$time_pso<-rep("never_inf", length(dta_01s$pid))
rm(tb1, tb2, X1_pone)
```

@@

# correct var names
colnames(dta_03s)<-c("roche_n", "abbott_n",
                     colnames(dta_03s[,3:6]))
rm(X3_mmc2)
```

```{r}
## Data-04, 2/8 months data PSO, extracted from Y. Nagakama et al.
X4_Nakagama_all <- read_excel("../1_data/private/Prob Model/4-anonymized_rocheabbott.xlsx",
                              col_types = c("numeric", "text", "numeric", 
                                            "numeric", "numeric", "numeric"))

dta_04<-X4_Nakagama_all

# correct col names
colnames(dta_04)<-c("age","gender","month_pso","abbott_n","roche_n","pid")
dta_04$pso <- dta_04$month_pso*30
hist(dta_04$pso, breaks = seq(0,540, by=30)) # not so much samples within 3-9 months

dta_04$time_pso[dta_04$pso<=90]<-"less_3_month"
dta_04$time_pso[dta_04$pso<=180 & dta_04$pso>90]<-"3-6_month"
dta_04$time_pso[dta_04$pso<=270 & dta_04$pso>180]<-"6-9_month"
dta_04$time_pso[dta_04$pso>270]<-"more_9_month"
table(dta_04$time_pso, useNA = 'ifany')

# All participants been sampled only once, wt=1
dta_04$sample_wt<-rep(1, length(dta_04$pid))
dta_04$study<-rep(4, length(dta_04$pid))

# clean the unused data
rm(X4_Nakagama_all)
```

```{r}
## Data-06, shared by Dr Paul Bieniasz
X6_Paul <- read_excel("../1_data/private/Prob Model/6-2024_02_06_SR1407_McGill.xlsx", 
    sheet = "data_import", skip = 1)
dta_06<-select(X6_Paul, 1:6, 9)
colnames(dta_06)<-c("age", "gender", "pid", "pso",
                    "abbott_n", "roche_n", "study")
dta_06s<-dta_06[!(is.na(dta_06$abbott_n) &
                  is.na(dta_06$roche_n)),]

# add time_pso based on time
dta_06s$time_pso[dta_06s$pso<=90]<-"less_3_month"
dta_06s$time_pso[dta_06s$pso<=180 & dta_06s$pso>90]<-"3-6_month"
dta_06s$time_pso[dta_06s$pso<=270 & dta_06s$pso>180]<-"6-9_month"
dta_06s$time_pso[dta_06s$pso>270]<-"more_9_month"
table(dta_06s$time_pso, useNA = 'ifany')
hist(dta_06s$pso, breaks = seq(0,540, by=30)) # some data from month 3-9

# add weights
tb1<-as.data.frame(table(dta_06s$pid))
colnames(tb1)<-c("pid","freq")
tb1$sample_wt<-1/tb1$freq
dta_06s<-merge(dta_06s, select(tb1, 1,3), by="pid",all.x = T, all.y = F)
rm(tb1, tb2, X6_Paul)
```

@@

```{r}
dta_all<-dplyr::bind_rows(dta_01s, dta_03s, 
                          select(dta_04, -c("month_pso")), dta_06s)

# add log quant assay (log_e transformation)
dta_all$log_roche_n<-log(dta_all$roche_n)    # COI for log_n = 0
dta_all$log_abbott_n<-log(dta_all$abbott_n)  # COI for log_n = 0.3365 (COI = 1.4)

# clean the gender var
table(dta_all$gender)
dta_all$gender[dta_all$gender %in% c("m")]<-"M"
dta_all$gender[dta_all$gender %in% c("f")]<-"F"

# save the rbinded harmonized data
save(dta_all, file="../1_data/private/Prob Model/dta_all.RData")

# summary analysis
hist(dta_all$pso, breaks = seq(0,540, by=30))
table(dta_all$time_pso, useNA = 'ifany')

# remove all intermediate data
rm(dta_01, dta_01s, dta_03, dta_03s, dta_04, dta_06, dta_06s)
```

@@

#### 1.1 Multinomial Regression Model:

```{r}
# Model 01: simple model, unweighted
mnlm01<-multinom(relevel(as.factor(time_pso), ref = "never_inf") ~ roche_n, data = dta_all)
summary(mnlm01)

@@

# Model 02: takes sample weights into account
mnlm02<-multinom(relevel(as.factor(time_pso), ref = "never_inf") ~ roche_n, 
                 data = dta_all, weights = sample_wt)
summary(mnlm02)
```

@@

```{r}
# Model 03  takes sample weights into account
mnlm03<-multinom(relevel(as.factor(time_pso), ref = "never_inf") ~ log_roche_n, 
                 data = dta_all, weights = sample_wt)
summary(mnlm03)
## use estimates/SE and 1.96 to calculate p-value manually, should be <0.001
```

##====================== Part II. Applying the model to Sero data =============##

#### 2.1 Model adjusting for age and sex

@@

#### 2.2 Apply Model to CBS Data

@@

CBS_prob_r<-as.data.frame(predict(mnlm03, newdata = CBS, "probs")) # num of >3 months = 0
mean(CBS_prob_r[,1], na.rm = T)  # never: 59.82%
mean(CBS_prob_r[,4], na.rm = T)  # <3 months: 25.31%
mean(CBS_prob_r[,2], na.rm = T)  # 3-6 month: 3.87%
mean(CBS_prob_r[,3], na.rm = T)  # 6-9 months: 2.98%
mean(CBS_prob_r[,5], na.rm = T)  # >9 months: 8.02%

CBS_prob_r$ever_inf <- 1-CBS_prob_r$never_inf
CBS_prob_r$less_6_month <- CBS_prob_r$less_3_month + CBS_prob_r$`3-6_month`
CBS_prob_r$less_9_month <- CBS_prob_r$less_6_month + CBS_prob_r$`6-9_month`

CBSs<-cbind.data.frame(select(CBS, pid, sampledate, sex, age, fsa,
                        province, roche_n, abbott_n, roche_s, log_roche_n),
                 CBS_prob_r[,c(4,6,7,8)])
```

```{r}
## Applying the model with age & sex
CBS$gender<-CBS$sex

CBS_prob_r2<-as.data.frame(predict(mnlm03.b, newdata = CBS, "probs")) # num of >3 months = 0
mean(CBS_prob_r2[,1], na.rm = T)  # 59.49%
mean(CBS_prob_r2[,2], na.rm = T)  # 32.45%
mean(CBS_prob_r2[,3], na.rm = T)  # 8.06%

@@

# Copy new vars to CBSs2
CBSs2$region<-CBSs$region
CBSs2$samplemonth<-CBSs$samplemonth
```

@@

```{r}
# create by-month estimates, by region
# create by-month est for each region
tb_all<-CBSs %>%
  group_by(samplemonth, region) %>%
  summarise(n = n() , 
            ever_inf= mean(ever_inf, na.rm = T), 
            less_3_month=mean(less_3_month, na.rm = T),
            less_6_month=mean(less_6_month, na.rm=T),
            less_9_month=mean(less_9_month, na.rm=T),
            )
# clean NAs
tb_all<-tb_all[!is.na(tb_all$region), ]
```

##===================== Part III. Exp: Reg Model Abbott N~Roche N =============##

This idea came from Alton, we can use the data we have for prob model training to generate another model: Abbott-N~Roche-N. Then apply the model to the CBS data and generate the sero-prevalence curve, see if it gets closer to the APL curve

```{r}
# The linear model using log_e transformed abbott_n and roche_n

# lm01, simple model
lm01<-glm(log_abbott_n ~ log_roche_n, 
          family = "gaussian",
          data = dta_all[!is.na(dta_all$age) &
                         !is.na(dta_all$gender),], weights = sample_wt)
summary(lm01) # AIC=1398

# lm02, add age and sex
lm02<-glm(log_abbott_n ~ log_roche_n + age + gender, 
          family = "gaussian",
          data = dta_all[!is.na(dta_all$age) &
                         !is.na(dta_all$gender),], weights = sample_wt)
summary(lm02) # AIC=1401.1, increased a bit

# lm02.a, add age only
lm02.a<-glm(log_abbott_n ~ log_roche_n + age, 
          family = "gaussian",
          data = dta_all[!is.na(dta_all$age) &
                         !is.na(dta_all$gender),], weights = sample_wt)
summary(lm02.a) # AIC=1400, increased a bit

# lm02.b, add sex only
lm02.b<-glm(log_abbott_n ~ log_roche_n + gender, 
          family = "gaussian",
          data = dta_all[!is.na(dta_all$age) &
                         !is.na(dta_all$gender),], weights = sample_wt)
summary(lm02.b) # AIC=1399.1, increased a bit

lrtest(lm02, lm01) # p=0.65, not so much diff

## Let's just use simple model lm01 then

# refit model 01, using those with missing age/sex info
lm01<-glm(log_abbott_n ~ log_roche_n, 
          family = "gaussian",
          data = dta_all, weights = sample_wt)
summary(lm01) 
```

#### Apply the model lm01 to the CBS data
```{r}
log_abbott_n<-predict(lm01, newdata = CBSs, "response")
summary(log_abbott_n)

CBSs$log_abbott_n <- log_abbott_n
CBSs$abbott_n_hat <- exp(CBSs$log_abbott_n)
summary(CBSs$abbott_n_hat)

abbott_n<-CBSs %>%
  select(abbott_n, abbott_n_hat) %>%
  .[is.na(.$abbott_n)==F,]

plot(abbott_n$abbott_n, abbott_n$abbott_n_hat)
# do a paired t-test
t.test(abbott_n$abbott_n, abbott_n$abbott_n_hat,
       paired = TRUE)
# paired t-test, log transformed
abbott_n$log_abbott_n<-log(abbott_n$abbott_n)
abbott_n$log_abbott_n_hat<-log(abbott_n$abbott_n_hat)
t.test(abbott_n$log_abbott_n, abbott_n$log_abbott_n_hat,
       paired = TRUE)

# Apply COI=1.4
CBSs$pos_abbott<-if_else(CBSs$abbott_n_hat<=1.4, 0, 1, missing = NA)
# Apply Roche COI=1.0
CBSs$pos_roche<-if_else(CBSs$roche_n<=1, 0, 1, missing = NA)
```

#### plot: sero-prevalence, using predicted Abbott positivity
```{r}
tb_abb<-CBSs %>%
  group_by(samplemonth, region) %>%
  summarise(n = n() , 
            pos_abbott= sum(pos_abbott, na.rm = T), 
            pos_roche= sum(pos_roche, na.rm = T)
            )
tb_abb$prop.abbott<-tb_abb$pos_abbott/tb_abb$n
tb_abb$prop.roche<-tb_abb$pos_roche/tb_abb$n

# clean NAs
tb_abb<-tb_abb[!is.na(tb_abb$region), ]

tb_abb_melt<-reshape2::melt(select(tb_abb, 1:2,6:7), 
                        id.vars = c("samplemonth", "region"),
                        variable.name = "assay",
                        value.name = "prop")

p03<-tb_abb_melt %>%
  ggplot(aes(x=samplemonth, y=prop, group=assay, colour=assay)) +
  geom_point(aes(colour=assay)) +
  geom_line(aes(colour=assay))  +
  theme_bw()

p03 + facet_wrap(~region)
```

@@

```{r}
# Reshape meltted data
tb_melt<-reshape2::melt(select(tb_all, 1:2,4:7), 
                        id.vars = c("samplemonth", "region"),
                        variable.name = "inf",
                        value.name = "prop")

@@

#### 2.3 Apply model to APL data

@@

#### 2.4 Apply model to CANPATH data
